Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-12 paper/paper.tex
Almost finished with pruned de Bruijn graphs.

% Graphs
\newcommand{\gindegree}{\ensuremath{\mathsf{in}}}
\newcommand{\goutdegree}{\ensuremath{\mathsf{out}}}
\newcommand{\glabel}{\ensuremath{\mathsf{label}}}
\newcommand{\gkey}{\ensuremath{\mathsf{key}}}
\newcommand{\gvalue}{\ensuremath{\mathsf{value}}}

@@

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$. We say that $v_{0}$ is the \emph{start} node and $v_{\abs{P}-1}$ is the \emph{end} node of the path. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$. We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ as we did with infinite character sequences in Section~\ref{sect:strings}.

@@

\begin{proof}
Let $P[0, k'-1] = v_{0} \dotsm v_{k'-1}$ be a prefix of path $P$, where $1 \le k' \le \min(k, \abs{P})$.
If $k' = 1$, we know that $G.\glabel(P[0]) = G.\glabel(v_{0})$ is a prefix of $G.\gkey(v_{0})$.
Now assume that $k' > 1$. By induction, we know that $G.\glabel(P[1, k'-1])$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{i}, v_{i+1}) \in E$ exists, $G.\glabel(v_{0}) \cdot G.\gkey(v_{1}) = G.\gkey(v_{0}) \cdot c$ (where $c \in \Sigma$), and hence $G.\glabel(P[0, k'-1])$ is a prefix of $G.\gkey(v_{0})$. As the first $\min(k, \abs{P})$ characters of $G.\gkey(v_{0})$ and $G.\glabel(P)$ are identical, the result follows.
\end{proof}

@@

The \emph{\kmer{k} index} is the simplest path index. In its most basic form, the \kmer{k} index consists of a parameter value $k$, a hash function $h$, and an array $A$ of \emph{key-value pairs} $(X, W)$, where $X \in \Sigma^{k}$ and $W \subseteq V$. Given a \kmer{k} $X \in \Sigma^{k}$, we start looking for key $X$ from $A[h(X)]$, and list the set of start nodes $W$ as matches if we find it. Searching using a hash table is fast, but we can only search for patterns of length $k$, and the array requires a lot of space.

Another representation of the \kmer{k} index trades query performance for the ability to search for shorter patterns. Instead of using a hash table, we sort the key-value pairs by their keys, and store the pairs in array $A$ in sorted order. Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, W_{i})$, where pattern $X$ is a prefix of the key $X_{i}$. We then list the union $\bigcup_{i=sp}^{ep} W_{i}$ as the set of matches.

As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. We now define the de~Bruijn graph of a graph for that purpose.

\begin{definition}[Maximal paths]
Let $G = (V, E)$ be a labeled graph. We say that path $P$ in the graph is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
\end{definition}

We can build a \kcollection{k} based on the labels of maximal paths by starting with the collection $\mathcal{S}$ of such labels. In order to transform $\mathcal{S}$ into a \kcollection{k}, we adjust the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence.

\begin{definition}[de~Bruijn graph of a graph]
Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be a \kcollection{k} of maximal path labels in $G$. Then the \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length. For each key-value pair $(X, W)$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = W$ to the node.

\begin{definition}[de~Bruijn graph as an index]
Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be its \orderk{k} de~Bruijn graph. Pattern $X \in \Sigma^{\ast}$ matches node $v \in V'$ if it prefix-matches string $G'.\gkey(v)$. The set of occurrences for pattern $X$ is the union of sets $G.\gvalue(v)$ for the matching nodes $v \in V'$.
\end{definition}

As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} (path labels that exist in the de~Bruijn graph but not in the input graph) with patterns of length at most $k$. With longer patterns, we have to \emph{verify} the results in graph $G$ to avoid false positives. We will develop the idea of using graphs similar to de~Bruijn graphs as path indexes further in the next section.

@@

De~Bruijn graphs are a special case of path graphs. They represent paths of length $k$ in the \emph{input graph} as nodes and paths overlapping on $k-1$ nodes of the input graph as edges. We can generalize the notion by allowing the nodes of the path graph to represent paths of different lengths.

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings containing $\#^{k}$ and $\$^{k}$. We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$. Then the \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
\begin{enumerate}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
\item each node has a key $G'.\gkey(v_{K}) = K$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{enumerate}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

% FIXME from here
- using path graphs as indexes
- k-index: has a node matching any k-path in the input graph
- edge condition: Y has predecessor with c -> edge (X, Y) if cY matches X
- k-equivalent: union of values for nodes matching X identical for all |X| <= k
- k-equivalent + edge condition: no false positives of length <= k; no false negatives

pruned de Bruijn graphs: merge nodes of DBG while keeping the path graph k-equivalent
similar to manifold de Bruijn graphs \cite{Lin2014}

>>>>

% Graphs
\newcommand{\gindegree}{\ensuremath{\mathsf{in}}}
\newcommand{\goutdegree}{\ensuremath{\mathsf{out}}}
\newcommand{\glabel}{\ensuremath{\mathsf{label}}}
\newcommand{\gkey}{\ensuremath{\mathsf{key}}}
\newcommand{\gvalue}{\ensuremath{\mathsf{value}}}
\newcommand{\gnode}{\ensuremath{\mathsf{node}}}

@@

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$. We say that $v_{0}$ is the \emph{start} node and $v_{\abs{P}-1}$ is the \emph{end} node of the path. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.

We generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ in a similar way as we did with infinite character sequences in Section~\ref{sect:strings}. We say that path $P$ is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.

@@

\begin{proof}
Let $P[0, k'-1] = v_{0} \dotsm v_{k'-1}$ be a prefix of path $P$, where $1 \le k' \le \min(k, \abs{P})$.
If $k' = 1$, we know that $G.\glabel(P[0]) = G.\glabel(v_{0})$ is a prefix of $G.\gkey(v_{0})$.

Now assume that $k' > 1$. By induction, we know that $G.\glabel(P[1, k'-1])$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{i}, v_{i+1}) \in E$ exists, $G.\glabel(v_{0}) \cdot G.\gkey(v_{1}) = G.\gkey(v_{0}) \cdot c$ (where $c \in \Sigma$), and hence $G.\glabel(P[0, k'-1])$ is a prefix of $G.\gkey(v_{0})$. As the first $\min(k, \abs{P})$ characters of $G.\gkey(v_{0})$ and $G.\glabel(P)$ are identical, the result follows.
\end{proof}

@@

The \emph{\kmer{k} index} is the simplest path index. In its most basic form, the \kmer{k} index consists of a parameter value $k$, a hash function $h$, and an array $A$ of \emph{key-value pairs} $(X, V_{X})$, where $X \in \Sigma^{k}$ and $V_{X} \subseteq V$. Given a \kmer{k} $X \in \Sigma^{k}$, we start looking for key $X$ from $A[h(X)]$, and list the set of start nodes $V_{X}$ as matches if we find it. Searching using a hash table is fast, but we can only search for patterns of length $k$, and the array requires a lot of space.

Another representation of the \kmer{k} index trades query performance for the ability to search for shorter patterns. Instead of using a hash table, we sort the key-value pairs by their keys, and store the pairs in array $A$ in sorted order. Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, V_{X_{i}})$, where pattern $X$ is a prefix of the key $X_{i}$. We then list the union $\bigcup_{i=sp}^{ep} V_{X_{i}}$ as the set of matches.

As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. In order to define the de~Bruijn graph of a graph, we build a \kcollection{k} based on the collection $\mathcal{S}$ of the labels of the maximal paths in graph $G = (V, E)$. If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$. We then transform $\mathcal{S}$ into a \kcollection{k} by adjusting the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence. If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = (s, j)$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$.

\begin{definition}[de~Bruijn graph of a graph]
Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. Then the \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length. For each key-value pair $(X, V_{X})$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = V_{X}$ to the node. We say that a pattern matches node $v \in V'$ if it prefix-matches string $G.\gkey(v)$. As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} (path labels that exist in the de~Bruijn graph but not in the input graph) with patterns of length at most $k$. With longer patterns, we have to \emph{verify} the results in graph $G$ to avoid false positives. We will develop the idea of using graphs similar to de~Bruijn graphs as path indexes further in the next section.

@@

De~Bruijn graphs are a special case of path graphs. They represent the paths of length $k$ in the \emph{input graph} as nodes and the pairs of paths overlapping on $k-1$ nodes of the input graph as edges. We can generalize the notion by allowing the nodes of the path graph to represent paths of different lengths.

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings containing $\#^{k}$ and $\$^{k}$. We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$. Then the \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
\begin{enumerate}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
\item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i)$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{enumerate}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

\begin{definition}[Path graph as an index]
Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be the \orderk{k} path graph of $G$ with key set $\mathcal{K}$. Pattern $X \in \Sigma^{\ast}$ \emph{matches} node $v \in V'$ if it prefix-matches string $G'.\gkey(v)$. If $V_{X}$ is the set of nodes $v \in V'$ matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\gvalue(X) = G'.\gvalue(V_{X}) = \bigcup_{v \in V_{X}} G'.\gvalue(v)$.
\end{definition}

In the general case, queries using a path graph may result in false positives for patterns $X$ of length $\abs{X} > \abs{K}$, where $K$ is the shortest key in the key set $\mathcal{K}$. To avoid this, we restrict our attention to path graphs that are equivalent to de~Bruijn graphs.

\begin{definition}[Equivalent path graphs]
Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs. We say that graphs $G$ and $G'$ are \emph{equivalent}, if we have $G.\gvalue(X) = G'.\gvalue(X)$ for all patterns $X \in \Sigma^{\ast}$.
\end{definition}

\begin{definition}[Pruned de~Bruijn graph]
Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be the \orderk{k} path graph of $G$ with key set $\mathcal{K}$. Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is equivalent to the \orderk{k} de~Bruijn graph of $G$, and $\abs{K} \le k$ for all $K \in \mathcal{K}$.
\end{definition}

Pruned de~Bruijn graphs can be built by merging nodes of a de~Bruijn graph. They are similar to manifold de Bruijn graphs \cite{Lin2014}.

\begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, and let $V_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$. If $\abs{V_{K}} > 0$ and $G'.\gvalue(u) = G'.\gvalue(v)$ for all $u, v \in V_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$ after we remove $G'.\gkey(v)$ for all $v \in V_{K}$ from the key set and replace them with $K$.
\end{lemma}

\begin{proof}
Let $X \in \Sigma^{\ast}$ be a pattern, and let $V_{X} \subseteq V'$ be the set of nodes matching it. If the pattern does not prefix-match string $K$, we have $V_{X} \cap V_{K} = \emptyset$, and the changes in the key set do not affect the set of occurrences for pattern $X$.

Otherwise let $G'' = (V'', E'')$ be the path graph with adjusted key set, and let $v_{K} \in V''$ be the node corresponding to key $K$. Then
\begin{align*}
G''.\gvalue(X) & = G''.\gvalue(V_{X} \setminus V_{K}) \cup G''.\gvalue(v_{K}) \\
 & = G'.\gvalue(V_{X} \setminus V_{K}) \cup G'.\gvalue(V_{X} \cap V_{K}) = G'.\gvalue(X).
\end{align*}
\end{proof}

\begin{lemma}[Correctness]\label{lemma:dbl-correct}
Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$. For every pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$, let $V_{X}$ be the set of start nodes of paths $P$ in graph $G$ with $G.\glabel(P) = X$. Then $V_{X} \subseteq G'.\gvalue(X)$, and
$V_{X} = G'.\gvalue(X)$ if $\abs{X} \le k$.
\end{lemma}

% FIXME
% Introduce the de Bruijn graph
% Set of nodes of DBG matching the pattern -> set of kmers in S prefix-matching the pattern
% -> set of start nodes in G for paths matching the first min(k, |X|) characters of the pattern.
% If |X| > k, this may be a superset of start nodes matching the entire pattern.
% Pruned DBG equivalent to DBG -> holds with G' as well.
\begin{proof}
Foo
\end{proof}
